---
title: "Lecture 17: Intro to R and Descriptive Statistics"
subtitle: "Ling 250/450: Data Science for Linguistics"
date: "Spring 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Welcome to the second half of the course! We're switching gears from Python and
text data to **R and quantitative data analysis**. The focus for the rest of the
semester:

- **Statistics and probability** in R
- **Manipulating and visualizing data** with the tidyverse and ggplot
- **Hypothesis testing**, applied to your term projects

This notebook covers two things at once: getting oriented to R as a language,
and understanding the foundational concepts of descriptive statistics.

---

## 0. RStudio Tour

### The Four Panes

RStudio organizes your workspace into four panes. When you first open it (with
no file open), you'll see three:

**R Console** (left): the interactive R interpreter — like the Python REPL you
already know. You can type R expressions here and see results immediately.

**Environment** (top-right): shows every variable you've defined in the current
session. As you create variables, they'll appear here with their values.

**Files / Plots / Packages / Help** (bottom-right): a multi-tab panel for
navigating your filesystem, viewing plots, managing installed packages, and
reading documentation.

### Opening This Notebook

1. In RStudio, go to **File → Open File** and navigate to this `.Rmd` file
2. The **Script/Notebook editor** opens as a fourth pane above the console
3. You'll see the document rendered as a mix of text and code chunks

### Running Code

- To run a single chunk: click the **Run** button (▶) at the top-right of
  the chunk, or press **Ctrl+Shift+Enter** (Windows/Linux) /
  **Cmd+Shift+Enter** (Mac)
- To run just the current line or a selection: **Ctrl+Enter** / **Cmd+Enter**
- Output and plots appear inline below each chunk

### Rendering (Knitting) This Document

Click the **Knit** button at the top of the editor pane to render the full
notebook to HTML. RStudio will run all chunks from scratch and open the output
in a new window (or the Viewer pane).

Before knitting, make sure you've installed `dplyr` (covered in §3) — the
notebook uses it, and knitting will fail if it's missing.

If knitting fails with an error about `rmarkdown` or `knitr` not being found,
run these once in the console:

```r
install.packages("rmarkdown")
install.packages("knitr")
```

These are usually pre-installed with RStudio, but may be missing on some systems.

---

## 1. What R Is Good For

R was designed specifically for data analysis and statistics. Its strengths:

- **Tabular data** (CSVs, data frames) — via the `tidyverse` library ecosystem
- **Statistical analysis** — many statistical tests are built in or easily installed
- **Data visualization** — R's `ggplot2` library is the gold standard for
  scientific figures

What R is **not** good for:

- **Text processing** — string operations are less intuitive than Python
- **General-purpose programming** — complex algorithms are cumbersome; Python is
  far more versatile

For this course: Python for text processing, R for statistics and visualization.

---

## 2. R Language Basics

### Similarities to Python

**Arithmetic operators** work the same way:

```{r arithmetic}
# Addition, subtraction
3 + 5
10 - 4

# Multiplication, division, exponentiation
6 * 7
20 / 4
2 ^ 8

# Comparisons
10 > 5
3 == 3
4 != 7
```

**Strings** are written in double quotes:

```{r strings}
my_string = "This is a string"
my_string
```

Note: In R, strings and individual characters are the same type, both called
"character" (unlike Python, which distinguishes `str` from `chr`). String
manipulation functions exist but are less intuitive than Python's.

**Comments** use `#`:

```{r comments}
# This is a comment — R ignores it
x = 42  # inline comments work too
```

### Minor Syntax Differences

**Variable assignment** — R has two equivalent options:

```{r assignment}
# Both of these do the same thing
my_variable = 10
my_variable <- 10

my_variable
```

The arrow operator (`<-`) is the traditional R style; `=` was added later and
is now widely accepted. We'll use `=` in this course.

**Boolean operators** — same logic as Python, different symbols:

```{r booleans}
# "and": & (not "and")
# "or": | (not "or")
# "not": ! (not "not")
# Boolean values: TRUE / FALSE (not True / False)

TRUE & FALSE
TRUE | FALSE
!TRUE

# Parentheses needed for order of operations
(2 > 1) & (3 > 2)
```

**Other constructs** (functions, if/else, loops) work similarly to Python but
with syntactic differences. We won't focus on these heavily — see *Learning
Statistics with R* Ch. 3 if you need them.

### Vectors: R's Version of Python Lists

The most important data structure difference between R and Python is how
sequences work. In Python, you use lists. In R, the equivalent is a **vector**.

```{r vectors}
# Create a vector with the c() function ("combine")
my_vec = c(1, 2, 3, 4, 5)
my_vec
```

Key differences from Python lists:

**1. Indexing starts at 1 (not 0):**

```{r indexing}
my_vec[1]   # first element is 1, not 0
my_vec[5]   # last element
```

**2. Elements can be reassigned:**

```{r element_reassign}
my_vec[1] = 0
my_vec
```

**3. Arithmetic applies to all elements at once (vectorized):**

```{r vectorized}
my_vec = c(1, 2, 3, 4, 5)  # reset

my_vec * 2         # multiply every element by 2
my_vec + c(1, 2, 3, 4, 5)  # element-wise addition of two vectors
```

**4. `length()` gives the number of elements:**

```{r length}
length(my_vec)
```

<details>
<summary>**Challenge:** What does `my_vec ^ 2` produce? Try it.</summary>

```{r challenge_vectors}
my_vec ^ 2
```

It squares each element. Vectorized operations apply element-wise.
</details>

### Data Frames

A **data frame** stores tabular data — rows and columns, like a spreadsheet or
CSV. This is R's equivalent of a pandas DataFrame (but we'll use R's built-in
and tidyverse tools instead).

```{r read_csv}
# Read a CSV file into a data frame
vowels = read.csv("../data/vowels.csv")

# View the first few rows
head(vowels)
```

You can also create a data frame from vectors:

```{r create_df}
evens = c(2, 4, 6, 8)
odds = c(1, 3, 5, 7)
data.frame(evens, odds)
```

**Accessing a column** with the `$` operator:

```{r dollar_operator}
# Get the VOWEL column as a vector
head(vowels$VOWEL)

# Get all column names
names(vowels)
```

---

## 3. Packages and dplyr

### Installing Packages

R packages extend the base language with additional functions. The main
repository for R packages is CRAN (the Comprehensive R Archive Network).

**Installing via RStudio's GUI** (recommended for new packages):

1. Go to the **Packages** tab in the bottom-right pane
2. Click **Install**
3. Type the package name (e.g., `dplyr`) — autocomplete will suggest matches
4. Make sure **Install dependencies** is checked, then click **Install**

You only need to install a package once. RStudio will show its progress in the
console.

**Alternatively**, you can install from the console directly:

```r
# Run in the console, not in a notebook chunk — you only need to do this once
install.packages("dplyr")
```

### Loading a Package

Once installed, load a package with `library()` (similar to Python's `import`):

```{r load_dplyr}
library(dplyr)
```

Unlike Python, you do **not** need to prefix function names with the package
name. After `library(dplyr)`, you call `filter()` directly — not `dplyr::filter()`.

### Filtering Data Frames with `filter()`

`filter()` returns a **subset** of a data frame where certain conditions hold:

```{r filter_basic}
# All rows where the vowel is "i"
filter(vowels, VOWEL == "i")
```

`filter()` can take **any number of conditions**, all of which must be true:

```{r filter_multi}
# Female speakers producing the vowels "i" or "u",
# whose height is at or below the mean height of all speakers
filter(
  vowels,
  SEX == "female",
  VOWEL == "i" | VOWEL == "u",
  HEIGHT <= mean(HEIGHT)
)
```

Note: `mean(HEIGHT)` here refers to the HEIGHT column of the data frame —
dplyr makes columns available by name within filter conditions.

<details>
<summary>**Challenge:** Filter for male speakers who produced the vowel "æ"
(written as "ae" in the data). How many rows are there?</summary>

```{r challenge_filter}
ae_males = filter(vowels, SEX == "male", VOWEL == "ae")
nrow(ae_males)
```
</details>

---

## 4. Variable Types and Scales of Measurement

### What Is a Variable?

A **variable** is a measurable attribute of data — something you record for
each observation. Variables come from:

- **Instrumentation or observation** (e.g., vowel formants measured with a
  microphone and audio software)
- **Human responses** (e.g., a survey answer)

Variables are usually stored in **columns** of a data frame or CSV. Critically,
variables are **not all on the same scale** — you can't directly compare birth
year, eye color, and vowel formants.

### Four Scales of Measurement

Our textbook (LSR) defines four scales. The scale tells you **how you should
analyze** a variable.

#### Nominal (Categorical)

Nominal variables have **no meaningful ordering or relationship** between values.
The only thing you can say is that values are different.

- Examples: eye color, sex, citizenship, language spoken, transportation method
- Neither French nor German is "greater than" the other, and they **can't be
  averaged**
- I will usually call these **categorical variables**

#### Ordinal

Ordinal variables have a **natural, meaningful ordering** between values, but
**no other structure**.

- Examples: position in a race, ranked choice voting, other rankings
- Finishing 1st vs. 2nd tells you who was faster, but **not by how much**
- Addition and subtraction are **not well-defined** (1st place isn't "3rd minus
  2nd")
- Ordinal values also **can't be meaningfully averaged**

#### Interval

Interval variables are **ordered** and **differences are meaningful**, but they
have **no true zero value**.

- Examples: Fahrenheit and Celsius, calendar year
- The difference between 60° and 65° is the same as between 20° and 25°
- But: 0°C is not "no heat," and 40° is not "twice as hot" as 20°
- **Not meaningful to multiply or divide**

#### Ratio

Ratio variables have **ordering**, **meaningful differences**, and a **true
zero**.

- Examples: race finish time, weight, age, sports scores
- A 30 y/o is 5 years older than a 25 y/o (30 − 25), or 1.2× older (30 / 25)
- Ratio variables **can be meaningfully averaged**
- These are what you might think of as "regular" numbers

### Continuous vs. Discrete

- **Continuous**: for any two values, it's possible to have a value in-between
  (e.g., between 2.0 and 3.0 is 2.5, then 2.05, etc.)
- **Discrete**: not continuous (e.g., the number that comes up after rolling a die)

| | continuous | discrete |
|---------|:---:|:---:|
| nominal | | ✓ |
| ordinal | | ✓ |
| interval | ✓ | ✓ |
| ratio | ✓ | ✓ |

*From LSR p.17*

Nominal and ordinal variables are always discrete. Interval and ratio variables
can be either.

### Exercise: Measurement Scales in the Vowels Dataset

The `vowels` data frame has these columns: SPEAKER, WORD, VOWEL, F1, F2, SEX,
HEIGHT.

What is the **measurement scale** of each variable?

<details>
<summary>**Answers**</summary>

| Column | Scale | Notes |
|--------|-------|-------|
| SPEAKER | nominal/categorical | just labels for different speakers |
| WORD | nominal/categorical | no ordering between words |
| VOWEL | nominal/categorical | for our purposes (vowel categories) |
| F1 | ratio | frequency in Hz; has a true zero |
| F2 | ratio | frequency in Hz; has a true zero |
| SEX | nominal/categorical | no ordering |
| HEIGHT | ratio | height in cm; has a true zero |

</details>

### Factors in R

Sometimes a **categorical variable is coded as a number** in a CSV — for
example, a "sex" column might use 0 for male, 1 for female. R will naively
treat these as numeric, which can cause problems:

```{r factor_problem}
sex_numeric = c(0, 1, 0, 0, 1, 0, 0, 0)

# R treats this as a number — inappropriate operations succeed silently
sex_numeric < 2       # all TRUE (meaningless)
sex_numeric / 8       # silently produces garbage
```

The R data type **factor** designates a categorical variable and prevents
inappropriate manipulation:

```{r factor_fix}
sex_factor = as.factor(sex_numeric)
sex_factor            # shows levels

# Now arithmetic raises a warning and returns NA
sex_factor + 1
```

You can give the **levels** meaningful names with `levels()`:

```{r factor_levels}
sex_named = as.factor(c(0, 1, 0, 0, 1, 0, 0, 0))
levels(sex_named) = c("male", "female")
sex_named
```

This also prevents accidental equivalence between two different factor
variables that happen to use the same numbers — after naming, R will error if
you try to compare factors with different level sets.

---

## 5. Descriptive Statistics

**Descriptive statistics** aim to **concisely and meaningfully summarize** data.
(As opposed to *inferential* statistics, which make inferences from data — more
on that in the coming weeks.)

Two broad categories:

- **Central tendency**: where the "middle" of the data is (mean, median, mode)
- **Variability**: how much the data deviates from the central tendency
  (variance, standard deviation)

### Central Tendency

#### Mean

The **mean** (what most people call the "average") is the sum of all data
points divided by the total number of data points:

$$\bar{X} = \frac{1}{N} \sum_{i=1}^{N} X_i$$

```{r mean}
mean(c(1, 2, 3, 4, 5))

# Mean F1 in the vowels dataset
mean(vowels$F1)
```

#### Median

The **median** is the middle datapoint when values are arranged in order. It is
equivalent to the **50th percentile**: half the data falls above it, half below.

```{r median}
# 3 is the median of both of these:
median(c(1, 2, 3, 4, 5))
median(c(1, 2, 3, 8, 10))   # notice mean would differ, but median is same

# For an even number of points, median = average of the middle two
median(c(1, 2, 3, 4))       # (2 + 3) / 2 = 2.5

# Median height in the vowels dataset
median(vowels$HEIGHT)
```

#### Mean vs. Median

**Advantages of the mean:**

- Takes every datapoint into account (median ignores all but the middle)
- Mathematically important — used in many statistical formulas
- Almost everyone understands it as the "average"

**Advantages of the median:**

- **Resistant to outlier datapoints** — this is why people report "median
  salary" or "median home price": a handful of ultra-high earners skew the
  mean upward
- Appropriate for **ordinal data**, whereas the mean is not

```{r mean_vs_median}
# Simulated salaries: most people earn ~50k, one billionaire
salaries = c(45000, 48000, 52000, 55000, 60000, 1000000000)
mean(salaries)
median(salaries)
```

### Variability

#### Variance

**Variance** is the mean squared deviation from the mean:

$$Var(X) = \frac{1}{N} \sum_{i=1}^{N} (X_i - \bar{X})^2$$

Squaring serves the same role as absolute value — it makes all deviations
positive. Variance is mathematically important, but the units are squared
(e.g., Hz²), which makes it hard to interpret directly.

```{r variance}
var(vowels$F1)
```

#### Standard Deviation

**Standard deviation** is the square root of the variance:

$$StDev(X) = \sqrt{\frac{1}{N} \sum_{i=1}^{N} (X_i - \bar{X})^2}$$

Often written as σ (sigma). Because it's in the **same units as the original
variable**, it's easier to interpret than variance. You'll often see a
distribution described as $\bar{X} \pm StDev(X)$.

```{r sd}
sd(vowels$F1)

# Report mean ± SD
cat("F1:", round(mean(vowels$F1), 1), "±", round(sd(vowels$F1), 1), "Hz\n")
```

#### Quantiles and Percentiles

A **quantile** (or **percentile**) is a value below which a certain percentage
of the data falls. The median is the 50th percentile.

```{r quantiles}
# 50th percentile of height (should equal the median)
quantile(vowels$HEIGHT, probs = 0.5)

# 25th, 50th, and 75th percentiles
quantile(vowels$HEIGHT, probs = c(0.25, 0.5, 0.75))
```

#### Interquartile Range (IQR)

The **IQR** is the difference between the 75th and 25th percentiles — it shows
how wide the **middle 50%** of the data is. Useful compared to the total range:

```{r iqr}
IQR(vowels$HEIGHT)

# Compare to total range
total_range = max(vowels$HEIGHT) - min(vowels$HEIGHT)
total_range
```

The IQR is 12.25 cm while the total range is 36 cm — the middle half of the
data is relatively compact compared to the full spread.

### Summary Statistics

R's `summary()` function gives several statistics in one call, and works on
either a vector or a full data frame:

```{r summary_vector}
summary(vowels$HEIGHT)
```

```{r summary_df}
summary(vowels)
```

Note that for character columns (SPEAKER, WORD, VOWEL, SEX), `summary()` just
reports the class and length — it can only compute numeric statistics for
numeric columns.

---

## 6. Basic Plotting

R has built-in functions for quick data visualization. These are great for
**exploratory analysis** — getting a fast look at your data. For
publication-quality figures, we'll use `ggplot2` (coming in a few weeks).

### Scatterplots with `plot()`

`plot()` with **one vector** plots the values on the y-axis (by index). With
**two vectors**, it makes a **scatterplot**:

```{r scatterplot}
# Scatterplot of F1 vs. F2
plot(vowels$F1, vowels$F2)
```

Notice the negative correlation — higher F1 tends to go with lower F2, which
corresponds to the vowel space in phonetics.

### Histograms with `hist()`

`hist()` plots the **frequency** of values in certain ranges, giving you a
sense of the data's distribution:

```{r histogram}
hist(vowels$HEIGHT)
```

The `breaks` argument controls approximately how many bins to use:

```{r histogram_breaks}
# Fewer bins — broader picture
hist(vowels$HEIGHT, breaks = 4)

# More bins — finer detail
hist(vowels$HEIGHT, breaks = 16)
```

<details>
<summary>**Challenge:** Plot a histogram of F1 values. What shape does the
distribution have?</summary>

```{r challenge_hist}
hist(vowels$F1, breaks = 20, main = "F1 Distribution", xlab = "F1 (Hz)")
```

The distribution is right-skewed — most F1 values are in the lower range, with
a long tail toward higher frequencies.
</details>

### Boxplots with `boxplot()`

`boxplot()` shows the **inner quartiles** of the data. The box covers the
middle 50% (from 25th to 75th percentile), the line inside is the median, and
the whiskers extend to the data's range (excluding outliers):

```{r boxplot}
boxplot(vowels$HEIGHT)
```

Note that `boxplot()` doesn't add axis titles by default. You can add them:

```{r boxplot_titled}
boxplot(vowels$HEIGHT, ylab = "Height (cm)", main = "Speaker Heights")
```

---

## Quick Reference

| Task | R code |
|------|--------|
| Create a vector | `c(1, 2, 3)` |
| Index a vector | `vec[1]` (1-based!) |
| Read a CSV | `read.csv("file.csv")` |
| Access a column | `df$COLUMN` |
| Get column names | `names(df)` |
| Load a package | `library(dplyr)` |
| Filter rows | `filter(df, condition)` |
| Convert to factor | `as.factor(vec)` |
| Name factor levels | `levels(fac) = c("a", "b")` |
| Mean | `mean(vec)` |
| Median | `median(vec)` |
| Variance | `var(vec)` |
| Standard deviation | `sd(vec)` |
| Percentiles | `quantile(vec, probs = 0.25)` |
| IQR | `IQR(vec)` |
| Summary stats | `summary(vec)` or `summary(df)` |
| Scatterplot | `plot(x_vec, y_vec)` |
| Histogram | `hist(vec, breaks = 10)` |
| Boxplot | `boxplot(vec)` |
